### 2017-02-17 real/draft.tex
[draft] Revamp tardis section

\subsection{Preprocessing \tardis{} output } \label{sec:tardis}

\hans{H: This subsection needs much more work. I have just copied some
  relevant text in here which may be useful. Much of it should be
  removed. Some connection to TARDIS should preferably be kept in the
  paper, in my view, eg the definition of $X = 1 - L/\ell_{\rm max}$.}

The \tardis{} packages simulates supernovae with %
\fred{Wolfgang's favorite description of his package}. %
Here we focus on the real packets as opposed to the virtual packets
output by tardis.

A histogram plot of the luminosities from a \tardis{} run indicates a
strongly skewed distribution with its single mode close to the maximum
luminosity $\lmax$, while there is a long tail for small $\ell$. These
features can be accomodated by a $\GammaDist$ distribution if we
transform the luminosities to
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly $X_j = 1 - (L_j/\lmax)$ and $\Lumtot \to X$. We assume
the total number of samples across all bins is large, hence the effect
of considering $\lmax$ a constant independent of the data is
negligible. We checked that leaving $\lmax$ as a free parameter in a
fit gave equivalent results.
%
Numerically, we take the largest luminosity of all \tardis{} packets
across the entire frequency range and increase it slightly to avoid
numerical difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_i \ell_i,
\end{align}
where $ \epsilon=10^{-6}$.

We infer $p(X \cond n,\bml)$ and only in the end transform back to a
probability density in $\Lumtot$ via
\begin{align}
  \lleq{gmg}
  p(\Lumtot  \cond n, \bml)
  &= p(X  \cond n, \bmx) \frac{1}{\lmax} \,.
\end{align}
To simplify the numerics, we also rescale the frequencies by a
constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  %

A major numerical simplification arises because of the $\Lumtot \to X$
transformation. Rewriting the \refeq{sll} formally as
\begin{align}
  \lleq{gmhb}
  p(Q  \cond n,\bml)
  & = \sum_{N} p(N \cond n)\; p(Q \cond N, n, \bml),
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single-packet luminosity is narrow
in the sense that the standard deviation relative to the mean is at
the percent level. This implies that $p(Q \cond N, n, \bml)$ has one
mode from each $N$ and all modes are clearly separated. The
transformation $\Lumtot \to X$ effects a strong overlap between the
distributions $p(X \cond N, n, \bmx)$ with similar $N$, so
$p(X \cond n, \bmx)$ can be well approximated by a unimodal
distribution.

We chose the Gamma distribution because it is \emph{stable}; i.e.,
the sum of $N$ Gamma variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
\end{align}
Assuming $\alpha = \alpha(\nu \cond \bmphi)$ and $\beta = \beta(\nu
\cond \bmphi)$ are functions of $\nu$ that depend on parameters
$\bmphi$, we set
\begin{align}
  \lleq{gmib}
  p(\bmX \cond \bmphi, N, \nu) &= \prod_{j=1}^N \GammaDist(X_j \cond \alpha, \beta ),\\
  p(\bmx \cond \bmphi, \npack, \bmnu) &= \prod_{i=1}^{\npack} \GammaDist(x_i \cond \alpha_i, \beta_i).
\end{align}
where $\alpha_i \equiv \alpha(\nu_i \cond \bmphi), \beta_i \equiv
\beta(\nu_i \cond \bmphi)$.

\subsection{Linear model}

\fred{Parts may still be relevant, especially if we put a plot of an
  entire spectrum in the paper.}

For simple functions $\alpha(\nu \cond \bmphi)$, one can find the
minimum value of $\alpha$ over the entire range of $\nu$
analytically. If this is too complicated, we propose to require
$\alpha(\nu_i \cond \bmphi) > 1$ only at each packet frequency
$\nu_i$. In the evaluation of $p(\bmx \cond \bmphi, \npack, \bmnu)$,
one has to loop over all packets and compute $\alpha$ and $\beta$
anyways, so the constraint imposes essentially no computational
overhead.

In the absence of a physical model to determine
$\alpha(\nu \cond \bmphi)$ and $\beta(\nu \cond \bmphi)$, we propose
to start with a simple polynomial ansatz
$\alpha(\nu \cond \bmphi) = \sum_{k=0}^\Kalpha \alpha_k \nu^k,
\beta(\nu \cond \bmphi) = \sum_{k=0}^\Kbeta \beta_k \nu^k$ such that
\begin{align}
  \lleq{gmp}
  \bmphi = (\alpha_0, \dots \alpha_\Kalpha , \beta_0, \dots \beta_\Kbeta)
\end{align}
The usual trade off between parsimony and the ability to accurately
model the frequency dependence determines the orders of the
polynomials, $\Kalpha$ and $\Kbeta$. This is outside the scope of this
paper. Here we choose $\Kalpha = 3$ and $\Kbeta = 1$.

@@

\subsection{Moments of the luminosity prediction}\label{sec:predict-moments}
\hans{H: I would agree to remove this section; I do not see value added.}

While the integral in \refeq{ral} cannot be integrated analytically,
the integrand is unimodal and strongly peaked, so that the Laplace or
saddle-point approximation~\cite[Ch. 27]{mackay2003information} can be
applied. If only the value $p(\Lumtot{=}\lumtot \cond n,m_1,m_2)$ at the
total data luminosity $q$ is sought, then one can approximate it
directly. In addition, we can compute the moments of $\Lumtot$ through
the \emph{moment generating function} \cite{stuart1994kendall}
\begin{align}
  \lleq{rap}
  M(t \cond \bmy) = \int_0^{\infty} \rmdx{\Lumtot} p(\Lumtot \cond \bmy) e^{-t \Lumtot} \,
\end{align}
from which the moments are accessible by differentiation
\begin{align}
  \lleq{raq}
  \expect{\Lumtot^k} = (-1)^k \eval{\dv[k]{M(t \cond \bmy)}{t}}_{t=0}
\end{align}
Assuming all involved integrals are absolutely convergent, we can
exchange the order of integration and find
\begin{align}
  \lleq{rar}
  M(t \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy)  f(t \cond \lambda, \mu, \sigma^2),
  \intertext{where }
  f\left(t \cond  \lambda, \mu, \sigma^2\right) &\equiv \int \rmdx{\Lumtot} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right) e^{-t \Lumtot}\\
  &= \frac{1}{2} \exp \qty(t/2 \lambda (t (\mu^2 + \sigma^2)- 2\mu)) \erfc \qty( \frac{\sqrt{\lambda}(t \qty(\mu^2 + \sigma^2) - \mu)}{\sqrt{2 \qty(\mu^2 + \sigma^2)}}).
\end{align}
This implies that the mean and variance of $\Lumtot$ require an integral over the three parameters
\begin{align}
  \lleq{ras}
  \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0}) - \expect{\Lumtot}^2
\end{align}
We implemented the above formulae in the julia language~\cite{julia14}
and used Newton's method from the \texttt{Optim.jl} package in
combination with automatic
differentiation~\cite{RevelsLubinPapamarkou2016} to optimize the
respective integrand and compute the Hessian at the mode. As initial
values, we slightly offset the sample estimates to avoid issues if the
optimizer cannot improve on the initial guess and found convergence
within $\order{10}$ steps:
\begin{align}
  \lleq{rat}
  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2).
\end{align}
\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods } %
\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %
\fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}

>>>>

\subsection{Preprocessing simulation output } \label{sec:tardis}

\fred{I removed a lot of cruft and swapped the symbols
  $x \to \ell, \ell \to \ell^\prime$ so it's in line with the previous section}

We consider \tardis{} as an example to demonstrate both how
transforming the output of a simulation helps in predictions, and what
further complications may arise in practice.

\tardis{} can output two kinds of packets, real and virtual. Here we
focus on the real packets and denote the raw output of \tardis{} by
primed symbols, for example for the $i$-{th} packet
$\qty(\nu_i^\prime, \ell_i^\prime)$.  A histogram plot
\fredmargin{include plot} of the luminosities from a \tardis{} run
indicates a strongly skewed distribution with its single mode close to
the maximum luminosity $\lmax^\prime$, while there is a long tail for
small $\ell$. These features can be accomodated by a $\GammaDist$
distribution as in \refsec{single-packet} if we transform the
luminosities to
\begin{align}
  \lleq{gmf}
  \ell_i \equiv 1- \frac{\ell_i^\prime}{\lmax^\prime} > 0
\end{align}
and similarly $L_i \equiv 1 - (L_i^\prime/\lmax^\prime)$. We assume
the total number of samples across all bins is large, hence the effect
of considering $\lmax^\prime$ a constant independent of the data is
negligible. We checked that leaving $\lmax^\prime$ as a free parameter
in a fit gave equivalent results.
%
Numerically, we take the largest luminosity of all \tardis{} packets
across the entire frequency range and increase it slightly to avoid
numerical difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax^\prime \equiv (1+\epsilon) \times \max_i \ell_i^\prime,
\end{align}
where $ \epsilon=10^{-6}$. We infer $p(\Lumtot \cond n,\bml)$ and only
in the end transform back to a probability density in the raw
coordinates via
\begin{align}
  \lleq{gmg}
  p(\Lumtot^\prime  \cond n, \bml^\prime)
  &= p(\Lumtot  \cond n, \bml) \frac{1}{\lmax^\prime} \,.
\end{align}
We also rescale the frequencies by a constant from
$\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$ but the most important
numerical simplification arises because of the transformation of the
luminosities. Rewriting \refeq{sll} formally as
\begin{align}
  \lleq{gmhb}
  p(Q  \cond n,\bml)
  & = \sum_{N} p(N \cond n)\; p(Q \cond N, n, \bml),
\end{align}
we see that the right-hand side is a mixture density. In the \tardis{}
application, the distribution of a single-packet luminosity,
$P\qty(\ell_i \cond \bmphi)$ is narrow in the sense that the standard
deviation relative to the mean is at the percent level. This implies
that each $p(Q^\prime \cond N, n, \bml^\prime)$ contributes one narrow
peak and all peaks are clearly separated in the superposition
$ p(Q^\prime \cond n,\bml^\prime)$. The transformation
$\ell^\prime \to \ell$ effects a strong overlap between the
distributions $p(\Lumtot \cond N, n, \bml)$ with similar $N$, so
$p(\Lumtot \cond n, \bml)$ can be well approximated by a unimodal
distribution and thus $\Lumtot$ instead of $\Lumtot^\prime$ the
preferred coordinate basis.
%
\fredmargin{show hedgehog plot}

\subsection{Towards an unbinned analysis} \label{sec:unbinned}

The parameters $\qty(\alpha, \beta)$ of the single-packet distribution
$\GammaDist(\ell_i \cond \alpha, \beta)$ may, of course, depend on the
frequency. % $\nu_i$
This effect could be captured, for example, by a (linear) model such
$\alpha = \alpha(\nu \cond \bmphi)$ and
$\beta = \beta(\nu \cond \bmphi)$ are functions of $\nu$ that depend
on parameters $\bmphi$ such that integrals over $\alpha, \beta$ would
turn into integrals over $\bmphi$. But for the sake of clarity, let us
consider only the simplest case where $\alpha, \beta$ are the same in
every bin such that \refeq{spk} becomes
\begin{align}
  \lleq{gmi}
  p(Q\cond \npack,\bmy)
  &=     \sum_{N=0}^\infty \;
    p(N\cond n)\;
    \int \rmdx{\alpha}\,\rmdx{\beta}\;
    \GammaDist(Q\cond N\alpha, \beta)\;
    p(\alpha, \beta \cond \npack, \bmy) \, .
\end{align}
In words, the posterior $ p(\alpha, \beta \cond \npack, \bmy)$ uses
all packets to essentially fix $\alpha, \beta$ if $\npack \gg 1$ and
allows meaningful predictions of $\Lumtot$ even in bins where
\emph{no} packet is observed; i.e., $n=0$. The only binning left is
performed to count $n$ the number of packets that fall into a given
frequency bin. This strategy is applicable, for example, when
absorption or emission lines in the spectrum abruptly alter the number
of packets for nearby frequencies but the luminosities are unaffected
by those line features. Note that if the dependence on $\bmphi$ can be
modeled with few parameters, the total number of parameters may be
drastically reduced: instead of having $\alpha, \beta$ separately for
each of a few thousand bins, there are potentially only a handful of
elements in $\bmphi$ for the entire spectrum. In that case, the
integral in \refeq{gmi} does not depend on the bin but only on $N$ so
its value may be cached to speed up the simultaneous evaluation across
all bins.

@@

\subsection{Moments of the luminosity prediction}\label{sec:predict-moments}
\fred{Shouldn't we use $n, \bml$ instead of $\bmy$ to be consistent?
  $\bmy$ would mean we use all samples from all bins}

While the integral in \refeq{ral} cannot be integrated analytically,
the integrand is unimodal and strongly peaked, so that the Laplace or
saddle-point approximation~\cite[Ch. 27]{mackay2003information} can be
applied. If only the value $p(\Lumtot{=}\lumtot \cond n,m_1,m_2)$ at the
total data luminosity $q$ is sought, then one can approximate it
directly. In addition, we can compute the moments of $\Lumtot$ through
the \emph{moment generating function} \cite{stuart1994kendall}
\begin{align}
  \lleq{rap}
  M(t \cond \bmy) = \int_0^{\infty} \rmdx{\Lumtot} p(\Lumtot \cond \bmy) e^{-t \Lumtot} \,
\end{align}
from which the moments are accessible by differentiation
\begin{align}
  \lleq{raq}
  \expect{\Lumtot^k} = (-1)^k \eval{\dv[k]{M(t \cond \bmy)}{t}}_{t=0}
\end{align}
Assuming all involved integrals are absolutely convergent, we can
exchange the order of integration and find
\begin{align}
  \lleq{rar}
  M(t \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy)  f(t \cond \lambda, \mu, \sigma^2),
  \intertext{where }
  f\left(t \cond  \lambda, \mu, \sigma^2\right) &\equiv \int \rmdx{\Lumtot} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right) e^{-t \Lumtot}\\
  &= \frac{1}{2} \exp \qty(t/2 \lambda (t (\mu^2 + \sigma^2)- 2\mu)) \erfc \qty( \frac{\sqrt{\lambda}(t \qty(\mu^2 + \sigma^2) - \mu)}{\sqrt{2 \qty(\mu^2 + \sigma^2)}}).
\end{align}
This implies that the mean and variance of $\Lumtot$ require an integral over the three parameters
\begin{align}
  \lleq{ras}
  \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty{\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0} - \qty(\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})^2} \, .
\end{align}
\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %
\fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}
